---
title: "lab16"
format: html
---

## Analyzing sequencing data in the cloud
Obtaining and processing SRA datasets on AWS


## Background
In bioinformatics we often need to analyze datasets that are too large (or would take too long) to analyze on our local lab computers. The goal of this hands-on session is to show you how to obtain and work with large biomolecular sequencing datasets using cloud computing resources. In particular, we will query, download, decompress and analyze data from NCBI’s main Sequence Read Archive (SRA) the worlds largest publicly available repository of high throughput sequencing data

## Importing Kalisto Results

```{r}
library(tximport)
library(readr)

# Get correct folders (only original `_quant`, not `_quant(1)`)
folders <- dir(pattern = "SRR[0-9]+_quant$", full.names = TRUE)

# Extract sample names
samples <- sub("_quant$", "", basename(folders))  

# Construct paths to abundance.tsv files instead of abundance.h5
files <- file.path(folders, "abundance.tsv")  
names(files) <- samples  

# Debugging: Verify paths
print(files)
print(file.exists(files))

# Read in tximport using abundance.tsv
txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

```{r}
library(tximport)
library(readr)

# Construct a temporary tx2gene mapping (transcripts as genes)
tx2gene <- data.frame(transcript_id = df$target_id, gene_id = df$target_id)

# Run tximport with abundance.tsv
txi.kallisto <- tximport(files, type = "kallisto", txOut = FALSE, tx2gene = tx2gene)

# Check structure
str(txi.kallisto)

```

```{r}
BiocManager::install("EnsDb.Hsapiens.v86")  # Modify for your organism
library(AnnotationDbi)
library(EnsDb.Hsapiens.v86)

# Generate tx2gene mapping
tx2gene <- transcripts(EnsDb.Hsapiens.v86, columns = c("tx_id", "gene_id"), return.type = "data.frame")

# Save for future use
write.csv(tx2gene, "tx2gene.csv", row.names = FALSE)

```

```{r}
# Install necessary packages if missing
if (!requireNamespace("BiocManager", quietly = TRUE)) 
    install.packages("BiocManager")

BiocManager::install(c("AnnotationDbi", "EnsDb.Hsapiens.v86", "tximport", "readr"))

# Load required libraries
library(tximport)
library(readr)
library(AnnotationDbi)
library(EnsDb.Hsapiens.v86)

# Generate transcript-to-gene mapping from Ensembl database
tx2gene <- transcripts(EnsDb.Hsapiens.v86, columns = c("tx_id", "gene_id"), return.type = "data.frame")

# Save tx2gene mapping for later use
write.csv(tx2gene, "tx2gene.csv", row.names = FALSE)

# Load tx2gene mapping
tx2gene <- read.csv("tx2gene.csv", header = TRUE)

# Get correct folders (only `_quant`, not `_quant(1)`)
folders <- dir(pattern = "SRR[0-9]+_quant$", full.names = TRUE)

# Extract sample names
samples <- sub("_quant$", "", basename(folders))  

# Construct paths to `abundance.tsv`
files <- file.path(folders, "abundance.tsv")  
names(files) <- samples  

# Debugging: Verify paths
print(files)
print(file.exists(files))

# Run tximport using tx2gene mapping
txi.kallisto <- tximport(files, type = "kallisto", txOut = FALSE, tx2gene = tx2gene)

# Check imported structure
str(txi.kallisto)

# View a preview of the imported counts
head(txi.kallisto$counts)

```

