---
title: "Class 12 Lab"
author: "Hugh (Trevor) Redford (PID:A17067426)"
date: 2/16/2025
format: pdf
---


# Propotion on G/G in a population

```{r}
# Read in Mexican Ancestry in LA csv file
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

> Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.Hint: The read.table(),summary() and boxplot() functions will likely be useful here. There is an example R script online to be used ONLY if you are struggling in vein. Note that you can find the medium value from saving the output of the boxplot() function to an R object and examining this object.There is also the medium() and summary() function that you can use to check your understanding.

The boxplot below displays the median expression levels corresponding to each genotype.

```{r}
#Read in file
gene_expression <-read.table("rs8067378_ENSG00000172057.6.txt")
head(gene_expression)
```

```{r}
nrow(gene_expression)
```

```{r}
table(gene_expression$geno)
```

```{r}
library(ggplot2)
ggplot(gene_expression) + aes(geno, exp, fill = geno) +
geom_boxplot(notch = TRUE)
```

> Q14. : Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need to be as polished as this one.

The boxplot shows ORMDL3 expression across genotypes (A/A, A/G, G/G). Individuals with A/A have higher expression than G/G, while A/G has the highest median. This suggests the SNP influences ORMDL3 expression, with the G allele linked to lower levels, highlighting its potential regulatory role.