---
title: "Class09 Halloween Mini Project"
author: "Hugh (Trevor) Redford PID:A17067426"
format: pdf
---

# Importing Candy Data
```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different candy types in the dataset.

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

There are 38 fruit candy types in the dataset.

# What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Reese's Peanut Butter cup and its win percent is 84.18029

```{r}
candy["Reese's Peanut Butter cup", ] $winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

The winpercent value for Kit Kat is 78.7686

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

The winpercent value for Tootsie Roll Snack Bars is 49.6535


# Skim Function

```{r}
library("skimr")
skim(candy)
```

> Q6.  Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the winpercent variable stands out as being on a significantly different scale than the other columns. While most variables have values ranging between 0 and 1, winpercent has a much higher mean (50.32), a larger standard deviation (14.71), and considerably greater percentile values. This suggests that winpercent is measured on a different numerical scale, making it an outlier in terms of magnitude compared to the other features in the dataset.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

It is likely that a value of 1 indicates that a candy contains chocolate, while a value of 0 signifies that the candy does not contain chocolate. This binary representation helps distinguish chocolate-based candies from non-chocolate ones in the dataset.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy) +
aes(winpercent) +
geom_histogram(bins = 10, col="red", fill="black")
```

> Q9. Is the distribution of winpercent values symmetrical?

No it is not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

It is below 50%.

> Q11.  On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# Chocolate Average Rank
chocolate.inds <- as.logical(candy$chocolate)
chocolate.wins <- candy[chocolate.inds,]$winpercent
mean(chocolate.wins)
```

```{r}
#Fruit Average Rank
fruit.inds <- as.logical(candy$fruity)
fruit.wins <- candy[fruit.inds,]$winpercent
mean(fruit.wins)
```

On average, chocolate candy (60.92153) is higher ranked than fruit candy (44.11974)

> Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.wins, fruit.wins)
```

The difference is statistically significant because the p-value from the Welch Two Sample t-test is 2.871e-08, which is far smaller than the commonly used significance threshold of 0.05. This indicates strong evidence against the null hypothesis, which assumes no difference between the mean wins of chocolate and fruit candies.

# Overall Candy Rankings

```{r}
# 5 least liked
head(candy[order(candy$winpercent),], n=5)
```

```{r}
# 5 most liked
tail(candy[order(candy$winpercent),], n=5)
```

> Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

> Q14. What are the top 5 all time favorite candy types out of this set?

Snickers, Kit Kat, Twix, Reese’s Miniatures, Reese’s Peanut Butter Cup

> Q15.  Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy)+
aes(winpercent, rownames(candy)) +
geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)
ggplot(candy)+
aes(winpercent, reorder(rownames(candy),winpercent)) +
geom_col()
```

```{r}
#Set color for each candy type
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
library(ggplot2)
ggplot(candy) +
aes(winpercent, reorder(rownames(candy),winpercent)) +
geom_col(fill=my_cols)
```

> Q17.  What is the worst ranked chocolate candy?

Nik L Nip

> Q18. What is the best ranked candy?

Reese's Peanut Butter Cup

# Taking a Look at Pricepercent

```{r}
library(ggrepel)
# How about a plot of price vs win
ggplot(candy) +
aes(winpercent, pricepercent, label=rownames(candy)) +
geom_point(col=my_cols) +
geom_text_repel(col=my_cols, size=3.3, max.overlaps = 50)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for
the least money - i.e. offers the most bang for your buck?

Reeses Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and
of these which is the least popular?

The top 5 most expensive candy types are Nik L Nip, Nestle Smarties, Hershey’s Krackel, Hershey’s Milk Chocolate, and Mr Good Bar. The least popular is Nik L Nip

> Q21. Make a barplot again with geom_col() this time using pricepercent
and then improve this step by step, first ordering the x-axis by value and finallymaking a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
aes(pricepercent, reorder(rownames(candy), pricepercent)) +
geom_segment(aes(yend = reorder(rownames(candy), pricepercent),
  xend = 0), col="gray40") +

geom_point()
```

# Exploring the Correlation Structure 

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and Chocolate

> Q23. Similarly, what two variables are most positively correlated?

Win Percent and Chocolate

# Principal Component Analysis (PCA)

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
#Main PCA score plot of PC1 vs. PC2
plot(pca$x[,1:2], col=my_cols, pch = 16)
```

```{r}
#Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) +
aes(x=PC1, y=PC2,
size=winpercent/100,
text=rownames(my_data),
label=rownames(my_data)) +
geom_point(col=my_cols)

print(p)
```

```{r}
library(ggrepel)
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 10) +
theme(legend.position = "none") +
labs(title="Halloween Candy PCA Space",
subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
caption="Data from 538")
```

> Q24. What original variables are picked up strongly by PC1 in the
positive direction? Do these make sense to you?

The primary variable strongly associated with PC1 in the positive direction is fruity. This makes sense because chocolate and fruity candies appear to be inversely correlated. Since chocolate-heavy candies tend to have negative PC1 values, it follows that fruity candies would be positioned toward the positive end of PC1. This suggests that PC1 effectively captures the contrast between chocolate-based and fruity candies in the dataset.

